---
title: "ESS523A_final"
author: "Rachel Hallmark"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Evaluating data from NEON sites simulated as C3 grasslands in the Community Land Model Version 5 

The purpose of this R markdown report is to compare data from two different NEON grassland sites. I will identify a variable of interest and then extract that variable from daily data collected every day for 4 years. Then I will plot the data over time.  

## Load Libraries 

Load libraries required for this project
```{r}

library(tidyverse)
library(ncdf4)
```

## Read in files and identify variables

Read in files for transient simulations from each site to identify variables of interest. 
```{r, results='hide'}
# open a file from one month for each site 

CPERjan2018 <- nc_open("simulation_data/CPER/CPER.C3mod/hist/CPER.C3mod.transient.clm2.h1.2018-01-01-00000.nc")
DSNYjan2018 <- nc_open("simulation_data/DSNY/DSNY.C3mod/hist/DSNY.C3mod.transient.clm2.h1.2018-01-01-00000.nc")

# view metadata - files are in netcdf format, so the metadata describes the data - want to identify names and units for variables of interest 
print(CPERjan2018)
print(DSNYjan2018)

```

## Explore GPP 

After viewing metadata, we want to look at the gross primary productivity variable, called "GPP" in units gC/m^2/s

```{r, results='hide'}
# name the variable as an object 
gpp <- "GPP"

# get gpp variable from both objects - this is an exploratory step to check and make sure we get a value. This is extracting data from a daily data file, so we should get 48 values - values are calculated every 30 minutes 
ncvar_get(CPERjan2018, gpp)
ncvar_get(DSNYjan2018, gpp) 



```

## Develop functions to get GPP from each daily data file for each site 

Now that we know which variable we want to explore, we need to get this variable from all of the daily data files. To do this - write two different functions, one to create a list of all the daily data files, and another to extract GPP from each daily data files. Then write these two functions into one single function that will do both tasks for each given site. 

```{r}


# write a function to make a list of all the files we want to get variables from 

site_list <- function(site) { 

  file_directory <- paste0(
   "simulation_data/",site,"/",site,".C3mod/hist"
   ) # points to directory 
  
  file_list <- list.files(file_directory, 
    pattern =   paste0(site,".C3mod.transient.clm2.h1.*.nc$"),     full.names = T) # creates a list of files with a defined pattern 

  return(file_list)
  
}

# write a function to extract and average the variable from each file 

get_avg_gpp <- function(y) {
  
  day <- nc_open(y) # opens a file and names it 'day'
    
  mean <- day %>% 
      ncvar_get(gpp) %>% 
    mean() # finds the mean of the gpp variable (defined earlier) for each 'day' file
  
    nc_close(day) 
    return(mean)
}

# put those two functions in a function that will perform both functions for each site given as an argument 

get_gpp <- 
  function(site) { 
    
  files <- site_list(site) 
  
  monthly_gpp <- map(files, get_avg_gpp)
  
  return(monthly_gpp)
 
  }

site <- c("CPER", "DSNY") # write sites into a character vector

sites_gpp <- map(site, get_gpp) # map over function for site list - returns a list with as many items that were provided as arguments, so 2 in this case 

names(sites_gpp) <- site # match site names with the names of list items

```

## Write function output into a data frame 

The output of that function gave a list, but we want these values in a data frame. 

```{r}
# write list into a data frame 
  # lists in sites_gpp are different lengths, so need to find the max length 
  max_length <- max(lengths(sites_gpp))
  # pad shorter list with NA values to match the longer list 
  sites_gpp_padded <-  map(sites_gpp, ~ c(.x, rep(NA, max_length - length(.x))))
  
# write lists into a dataframe 
  sites_gpp_df <- as_tibble(sites_gpp_padded)
```


## Make a list of days to match with daily data output 

Now we have a data frame with GPP values from each site, so we now need to match each value with the day it's from. 

```{r}
#make a list of days to match the number of rows in the data frame
Days <- seq(as.Date("2018-01-01"), by = "days", length.out = 1521)
head(Days) # check structure of the list 

# create a new column in sites_gpp_df for the day list 
sites_daily_gpp_df <- tibble(DAY = Days, sites_gpp_df)

# remove rows containing NA 
sites_daily_gpp_df <- na.omit(sites_daily_gpp_df)

#write into a csv 
write_csv(sites_daily_gpp_df, "CPER_DSNY_C3mod_GPP.csv")

```

## Tidy Data and Plot 

Now we have extracted the variable of interest from simulation data, lets visually compare the data from the two sites. 

```{r, warning=FALSE}

# pivot longer to get into correct format to plot 
sites_daily_gpp_long <- sites_daily_gpp_df %>%
  pivot_longer(cols = - DAY, 
               names_to = "site", 
               values_to = "Daily_GPP") %>% 
  mutate(Daily_GPP = unlist(Daily_GPP) %>% as.numeric())

head(sites_daily_gpp_long)

# write long data frame into a csv 
write_csv(sites_daily_gpp_long, "CPER_DSNY_C3mod_longDF.csv")


ggplot(data = sites_daily_gpp_long) + 
  geom_line(aes(x = DAY, y = Daily_GPP, colour = site)) + 
  labs(y = "daily gross primary productivity 
       (gC/m^2/s)", 
       x = "year", 
       title = "Simulated Daily GPP of NEON C3 grassland sites in CLM5") + 
  theme(plot.title = element_text(size = 10, hjust = 0.3),
        axis.title.y = element_text(size = 10))
  
```

It looks like GPP at the CPER site gets much lower than GPP at the DSNY site. This makes sense because the CPER site is located in Colorado, whereas the DSNY site is located in Florida. Grasses in Colorado may die off in the winter and not photosynthesize, whereas grasses in Florida are likely able to photosynthesize year round due to the milder climate. Despite being simulated as the same type of grass, the GPP between the two sites looks to have a different phenology. 





